RNeXML: a package for reading and writing richly annotated phylogenetic, character, and trait data in R

### 2015-01-08 manuscripts/manuscript.Rmd
break supplement into vignettes

---
#layout: preprint
layout: review, 11pt
title: "RNeXML: a package for reading and writing richly annotated phylogenetic, character, and trait data in R"
author: 
  - name: Carl Boettiger
    affiliation: cstar
    email: cboettig(at)gmail.com
    footnote: Corresponding author
  - name: Hilmar Lapp
    affiliation: NESCent
  - name: Scott Chamberlain
    affiliation: ropensci
  - name: Rutger Vos
    affiliation: NBC
address: 
  - code: cstar
    address: | 
      Center for Stock Assessment Research, 
      Department of Applied Math and Statistics, 
      University of California, Mail Stop SOE-2,
      Santa Cruz, CA 95064, USA
  - code: NESCent
    address: | 
      National Evolutionary Synthesis Center and Duke University, Durham, NC, USA
  - code: ropensci
    address: |
      University of California, Berkeley, CA, USA
  - code: NBC
    address: | 
      Naturalis Biodiversity Center, Leiden, the Netherlands
abstract: | 
      1. NeXML is a powerful and forward-compatible exchange standard
      recently proposed to better meet the expanding needs for
      machine-proof phylogenetic data and metadata sharing. Here we
      present the RNeXML package, which aims to provide to users of
      the popular R programming language easy programmatic access to
      reading and writing NeXML documents, including rich metadata, in
      a way that interfaces seamlessly with the extensive library of
      phylogenetic tools already available in the R ecosystem.

@@

Users of the popular statistical and mathematical computing platform R
[@R] enjoy a wealth of readily installable comparative phylogenetic
methods and tools <!-- needs a citation to support it -->. Exploiting
the opportunities arising from this wealth for complex and integrative
comparative research questions relies on the ability to reuse and
integrate previously generated or published data and metadata. The
expanding data exchange needs of the evolutionary research
community are rapidly outpacing the capabilities of most current
and widely used data exchange standards [@Vos_2012], which were all
developed a decade or more ago. This has resulted in a radiation of
different data representations and exchange standard "flavors" that
are no longer interoperable at the very time when the growth of
available data and methods has made that interoperability most
valuable. In response to the unmet needs for standardized data
exchange in phylogenetics, a modern XML-based exchange standard,
called NeXML, has recently been developed [@Vos_2012]. NeXML
comprehensively supports current data exchange needs, is predictably
machine-readable, and is forward compatible.

@@

Metadata annotatuons in `NeXML` can be nested within another
annotation, which the `meta()` method accommodates by accepting a
parameter `children`, with the list of nested metadata objects (which
can themselves be nested) as value.

@@

This project was supported in part by the National Evolutionary
Synthesis Center (NESCent) (NSF #EF-0905606), and grants from the
Natioal Science Foundation (DBI-1306697) and the Alfred P Sloan
Foundation (Grant 2013-6-22). `RNeXML` started as a project idea for
the Google Summer of Code(TM), and we thank Kseniia Shumelchyk for taking
the first steps to implement it. We are grateful to F. Michonneau for
helpful comments on an earlier version of this manuscript.

>>>>

---
#layout: preprint
layout: review, 11pt
title: "RNeXML: a package for reading and writing richly annotated phylogenetic, character, and trait data in R"
author: 
  - name: Carl Boettiger
    affiliation: cstar
    email: cboettig(at)gmail.com
    footnote: Corresponding author
  - name: Scott Chamberlain
    affiliation: ropensci
  - name: Rutger Vos
    affiliation: NBC
  - name: Hilmar Lapp
    affiliation: NESCent
address: 
  - code: cstar
    address: | 
      Center for Stock Assessment Research, 
      Department of Applied Math and Statistics, 
      University of California, Mail Stop SOE-2,
      Santa Cruz, CA 95064, USA
  - code: NESCent
    address: | 
      National Evolutionary Synthesis Center and Duke University, Durham, NC, USA
  - code: ropensci
    address: |
      University of California, Berkeley, CA, USA
  - code: NBC
    address: | 
      Naturalis Biodiversity Center, Leiden, the Netherlands
abstract: | 
      1. NeXML is a powerful and forward-compatible exchange standard
      recently proposed to better meet the expanding needs for
      machine-proof phylogenetic data and metadata sharing. Here we
      present the RNeXML package, which aims to provide to users of
      the popular R programming language easy programmatic access to
      reading and writing NeXML documents, including rich metadata, in
      a way that interfaces seamlessly with the extensive library of
      phylogenetic tools already available in the R ecosystem.

@@

Users of the popular statistical and mathematical computing platform
R [@R] enjoy a wealth of readily installable comparative phylogenetic
methods and tools [@taskview]. Exploiting the opportunities arising from
this wealth for complex and integrative comparative research questions
relies on the ability to reuse and integrate previously generated or
published data and metadata. The expanding data exchange needs of the
evolutionary research community are rapidly outpacing the capabilities of
most current and widely used data exchange standards [@Vos_2012], which
were all developed a decade or more ago. This has resulted in a radiation
of different data representations and exchange standard "flavors" that are
no longer interoperable at the very time when the growth of available data
and methods has made that interoperability most valuable. In response to
the unmet needs for standardized data exchange in phylogenetics, a modern
XML-based exchange standard, called NeXML, has recently been developed
[@Vos_2012]. NeXML comprehensively supports current data exchange needs,
is predictably machine-readable, and is forward compatible.

@@

Metadata annotations in `NeXML` can be nested within another
annotation, which the `meta()` method accommodates by accepting a
parameter `children`, with the list of nested metadata objects (which
can themselves be nested) as value.

@@

This project was supported in part by the National Evolutionary
Synthesis Center (NESCent) (NSF #EF-0905606), and grants from the
National Science Foundation (DBI-1306697) and the Alfred P Sloan
Foundation (Grant 2013-6-22). `RNeXML` started as a project idea for
the Google Summer of Code(TM), and we thank Kseniia Shumelchyk for taking
the first steps to implement it. We are grateful to F. Michonneau for
helpful comments on an earlier version of this manuscript.
